---
title: "ST104 catchup lab 9"
output: html_notebook
---

 **Anemia treatment study**
In this exercise, we look at one of the most common ways that the regression
fallacy can confuse results of an observational study. Suppose a company wants
to promote a nutritional supplement that they believe can increase red blood cell
count and therefore combat anemia. They recruit patients with low red blood cell
count, instruct them to eat the supplement for two weeks, and then measure their
red blood cell count again. The supplement is deemed to be a success if their red
blood cell count has increased on average (by more than would be predicted by
chance).
For this exercise you will first simulate data representing the whole population.
Then you will select the group the study recruited their patients from. Finally, you
will compare what can be observed from the whole population with what can be
observed on the selected population and draw conclusions.

Suppose the following facts:
 Each person has a “true blood cell count”, which is normally distributed and
does not change over time. For men this distribution has mean 150 and SD
6. For women this distribution has mean 133 and SD 5. All measurements
have the unit grams per litre, or g/L.
 Each measurement of red blood cell count adds a chance variation, which is
independent of the red blood cell count and is normally distributed with mean
0 and SD 2. This accounts for both the difficulty in precisely determining the
mass of red blood cells in a sample, and also for the fact that one sample of
blood from a patient may not contain the average red blood cell count for
that patient.
 In this study, each person’s red blood cell count is measured just twice, once
before and once after the supplement course.
 Male patients are judged to have anemia if their red blood cell count is below
130 g/L. Female patients are judged to have anemia if their red blood cell
count is below 120g/L. If a patient does not have anemia then they are
excluded from the study.
 The supplement has no effect. That is, the difference between the first and
second measurements is due only to chance variation, though the people
carrying out the study don’t know that.

1. Write a function rbccF() that inputs a number n and outputs an array with n
rows and 2 columns, representing the simulated measurements of red blood
cell count on n female patients. 



```{r}
rbccF <- function(n){
  #1st measurement
  truecount <- rnorm(n, 133, 5)
  m_1 <- truecount + rnorm(n,0,2)
  #2nd measurement
  m_2 <- truecount + rnorm(n,0,2)
  m <-cbind(m_1, m_2)
  return(m)
}
```
Write a function rbccM() that inputs a
number n and outputs an array with n rows and 2 columns, representing the
simulated measurements of red blood cell count on n male patients. 
```{r}
rbccM <- function(n){
  truecount <- rnorm(n, 150, 6)
  m_1 <- truecount + rnorm(n,0,2)
  m_2 <- truecount + rnorm(n,0,2)
  m <-cbind(m_1, m_2)
  return(m)
}
```
Combine these functions to create a function rbccT() that inputs a number n
and outputs an array with 2n rows and 2 columns, representing the simulated
measurements of red blood cell count on n female patients and n male patients, for 2n patients in total. In each case the first column should represent
the first test for red blood cell count, and the second column should represent
the second test for red blood cell count
```{r}
rbccT <- function(n){
  male <- rbccM(n)
  female <- rbccF(n)
  return(rbind(female, male))
}
```

Let B be the output of rbccT(1000).
Make a scatterplot of first and the second red blood cell count measurements
in B, with second measurements on the vertical axis. The plot should distinguish between male and female patients by e.g., colour, size or shape of the
points. You may find it helpful here to remember that the points() command
can be used to add new points into an existing plot.
```{r}
B <- rbccT(1000)
```
Make a scatterplot of first and the second red blood cell count measurements
in B, with second measurements on the vertical axis. The plot should distinguish between male and female patients by e.g., colour, size or shape of the points.

```{r}
plot(B[1:1000,1], B[1:1000,2], xlab="first measurement", ylab="second measurement", col="purple", xlim=c(min(B[,1]), max(B[,1])), ylim=c(min(B[,2]), max(B[,2])))
points(B[1001:2000,1], B[1001:2000,2], col="black")
# Add a legend to the graph
legend("bottomright", c("Male","Female"), pch=1, col = c("black","purple"),
inset = 0.02)
```
Note that some of these measurements in the second column of B will never
actually occur, because the associated individuals would have been excluded
from the study once they were found to not be anemic. 
Compute B1, the subarray of B of those subjects judged to have anemia, and so included in
the study. 
You could do this by directly removing both male and female
patients from B, or you could split B into two arrays, one with only male
patients and the other with only female, then remove non-anemic patients
from each of these new arrays before joining them together. Either way, you
may find it helpful to recall the syntax for subsetting arrays (e.g., define A<-
array(c(1 : 12), dim = c(3, 4)) and extract rows by A[A[, 2] < 6, ].) 

```{r}
#select eligible females
B1_F <- B[1:1000,][B[1:1000,1]<120,]
#select eligible males
B1_M <- B[1001:2000,][B[1001:2000,1]<130,]
#bind them together
B1 <- rbind(B1_F,B1_M)
```
By usingthe points() command, or by some other method, change the colour of the
points corresponding to subjects included in the study to red.
```{r}
plot(B[1:1000,1], B[1:1000,2], xlab="first measurement", ylab="second measurement", col="purple", xlim=c(min(B[,1]), max(B[,1])), ylim=c(min(B[,2]), max(B[,2])))
points(B[1001:2000,1], B[1001:2000,2], col="black")
points(B1[,1], B1[,2], col="red")
# Add a legend to the graph
legend("bottomright", c("Male","Female", "Anemic"), pch=1, col = c("black","purple", "red"),
inset = 0.02)
```
3. What is the correlation between the first and second measurements for all
rows of B (so both anemic and non-anemic patients)? Try repeating the
simulation many times, each time generating a new B and finding the correlation. Can you give an estimate of how much an estimate of correlation for a dataset like this would typically be off by, purely by chance?
There are several reasonable answers to the latter. You just need to give one
option. Think of a measure of spread.
```{r}
cor(B[,1],B[,2])
B <- rbccT(10000)
cor(B[,1],B[,2])
B <- rbccT(10000)
cor(B[,1],B[,2])
B <- rbccT(10000)
cor(B[,1],B[,2])
B <- rbccT(10000)
cor(B[,1],B[,2])
cor(B[,1],B[,2])
B <- rbccT(10000)
cor(B[,1],B[,2])
B <- rbccT(10000)
cor(B[,1],B[,2])
B <- rbccT(10000)
cor(B[,1],B[,2])
B <- rbccT(10000)
cor(B[,1],B[,2])
```
It would be off by about 33% COME BACK TO THIS

5. Use the lm() function to compute the regression line for predicting second
measurements from first measurements, based on the complete set of measurements B.
```{r}
Bln <- lm(B[,1]~B[,2])
B1ln <- lm(B1[,1]~B1[,2])
```
Put the line into the scatterplot. Repeat this for just the subjects included in the study, B1, plotting the new line in red. 
```{r}
plot(B[1:1000,1], B[1:1000,2], xlab="first measurement", ylab="second measurement", col="purple", xlim=c(min(B[,1]), max(B[,1])), ylim=c(min(B[,2]), max(B[,2])))
points(B[1001:2000,1], B[1001:2000,2], col="black")
points(B1[,1], B1[,2], col="red")
abline(Bln)
abline(B1ln, col="red")
```
Obviously, your results will depend on the simulated outputs. Draw three scatterplots and the
regression lines for the whole and the restricted data sets as above, each time
using a different pair of arrays B and B1.
```{r}
B <- rbccT(1000)
#select eligible females
B1_F <- B[1:1000,][B[1:1000,1]<120,]
#select eligible males
B1_M <- B[1001:2000,][B[1001:2000,1]<130,]
#bind them together
B1 <- rbind(B1_F,B1_M)

plot(B[1:1000,1], B[1:1000,2], xlab="first measurement", ylab="second measurement", col="purple", xlim=c(min(B[,1]), max(B[,1])), ylim=c(min(B[,2]), max(B[,2])))
points(B[1001:2000,1], B[1001:2000,2], col="black")
points(B1[,1], B1[,2], col="red")

Bln <- lm(B[,2]~B[,1])
B1ln <- lm(B1[,2]~B1[,1])

abline(Bln)
abline(B1ln, col="red")
```
```{r}
B <- rbccT(1000)
#select eligible females
B1_F <- B[1:1000,][B[1:1000,1]<120,]
#select eligible males
B1_M <- B[1001:2000,][B[1001:2000,1]<130,]
#bind them together
B1 <- rbind(B1_F,B1_M)

plot(B[1:1000,1], B[1:1000,2], xlab="first measurement", ylab="second measurement", col="purple", xlim=c(min(B[,1]), max(B[,1])), ylim=c(min(B[,2]), max(B[,2])))
points(B[1001:2000,1], B[1001:2000,2], col="black")
points(B1[,1], B1[,2], col="red")

Bln <- lm(B[,2]~B[,1])
B1ln <- lm(B1[,2]~B1[,1])

abline(Bln)
abline(B1ln, col="red")
```
```{r}
B <- rbccT(1000)
#select eligible females
B1_F <- B[1:1000,][B[1:1000,1]<120,]
#select eligible males
B1_M <- B[1001:2000,][B[1001:2000,1]<130,]
#bind them together
B1 <- rbind(B1_F,B1_M)

plot(B[1:1000,1], B[1:1000,2], xlab="first measurement", ylab="second measurement", col="purple", xlim=c(min(B[,1]), max(B[,1])), ylim=c(min(B[,2]), max(B[,2])))
points(B[1001:2000,1], B[1001:2000,2], col="black")
points(B1[,1], B1[,2], col="red")

Bln <- lm(B[,2]~B[,1])
B1ln <- lm(B1[,2]~B1[,1])

abline(Bln)
abline(B1ln, col="red")
```
What is the mean of the differences between the first and second measurements in B? In B1? 
```{r}
abs(sum(B[,1])/nrow(B)-sum(B[,2])/nrow(B))
abs(sum(B1[,1])/nrow(B1)-sum(B1[,2])/nrow(B1))
```
Again simulate many alternative versions of these datasets to gauge whether the mean of the differences could be purely due to chance.
```{r}
B_diff <- vector()
for (i in 1:25) {
  B <- rbccT(1000)
  B_diff <- c(B_diff,abs(sum(B[,1])/nrow(B)-sum(B[,2])/nrow(B)))
  }
print(B_diff)
```
```{r}
B1_diff <- vector()
for (i in 1:25){
  B <- rbccT(1000)
  #select eligible females
  B1_F <- B[1:1000,][B[1:1000,1]<120,]
  #select eligible males
  B1_M <- B[1001:2000,][B[1001:2000,1]<130,]
  #bind them together
  B1 <- rbind(B1_F,B1_M)
  B1_diff <-c(B1_diff, abs(sum(B1[,1])/nrow(B1)-sum(B1[,2])/nrow(B1)))
}
print(B1_diff)
```
The results don't seem to be purely due to chance. The differences between B1 results are much much bigger than the ones for B group


Consider the first twenty values for the mean of the differences between the first and second measurements of B. Consider the first twenty values for the mean of the differences between the first and second measurements of B1.
Explain the results in terms of the regression fallacy.
```{r}
B <- rbccT(10000)
  
#select eligible females
B1_F <- B[1:10000,][B[1:10000,1]<120,]
#select eligible males
B1_M <- B[1001:20000,][B[1001:20000,1]<130,]
#bind them together
B1 <- rbind(B1_F,B1_M)
  
abs(sum(B[1:50,1])-sum(B[1:50,2]))/50
abs(sum(B1[1:50,1])-sum(B1[1:50,2]))/50
```
The results from B1 are exceptionally low so it makes sense that repeating a trail (with tratment or without) yields a result closer to the mean and so with greater distance. Someone might incorrectly conclude that the treament helped whereas in reality it was the first measurement that was an outlier of sorts


