---
title: "Blood pressure"
author: "Dodzia"
output: html_notebook
---

A study was conducted into the effectiveness of a new blood pressure treatment at different doses. Patients’ blood pressures were measured initially, before 4 weeks of treatment with the new drug. Each patient was randomly allocated one of 4 different dose levels. At the end of the treatment period, the patients’ blood pressures were remeasured and the change was recorded. 

```{r}
blood <- read.csv("./data/pc3data1.csv")
```
Plot two graphs showing BPchange against BPinit and against Dose. 

```{r}
par(mfrow=c(2,1))
plot(BPchange ~ BPinit, data=blood, xlab="Initial BP (mmHg)", ylab="Change in BP (mmHg)")
plot(BPchange ~ Dose, data=blood, xlab='Dose (µg)', ylab="Change in BP (mmHg)")
```
The relation appears to be linear. So we fit it
```{r}
lm1 <- lm(BPchange ~ BPinit + Dose, data=blood)
summary(lm1)
```
The p-values are very small, advocating in favour of including Dose and BPinit in the model.
We carry out a sequential ANOVA by calling anova() with 1 argument
```{r}
anova(lm1)
```
Produce an ANOVA table for the test for the existence of regression and calculate the p-value. How do the quantities that you computed for this ANOVA table relate to the quantities in c?

The sum of squares for the ANOVA table is calculated as follows:
TotalSS is sum of entries in column "Sum Sq"
RegSS is the sum of the sum of squares for BPinit and Dose
ResidSS is stated explicitly above
```{r}
TotalSS <- 47.214+172.313+42.248
RegSS <- 47.214+172.313
ResidSS <- 42.248
```
a more elegant way to get this is
```{r}
a <- anova(lm1)
SS <- a[["Sum Sq"]] #the Sum Sq column
TotalSS <- sum(SS) # Total sum of squares
RegrSS <- sum(SS[1:2]) # Regression sum of squares
ResidSS <- SS[3] # Residual sum of squares
```
We now get the required degrees of freedom
```{r}
df <- a$Df
TotalDF <- sum(df) # Total degrees of freedom
RegrDF <- sum(df[1:2]) # Regression degrees of freedom
ResidDF <- df[3] # Residual degrees of freedom
```
We now compute the relevant mean squares and F-statistic







