---
title: "ST221 practical 1"
author: "Dodzia"
output: html_notebook
---

Consider the coursework marks (out of 20) and exam marks (out of 100) for some MORSE students

```{r}
cw <- c(5,5,10,12,14,16)
exam <- c(38,45,69,65,75,96)
```

Produce the first scatterplot
```{r}
plot(cw, exam, col = "purple", main = "cw vs exam marks", xlab = " coursework mark (out of 20)", ylab = "exam mark (out of 100)")
```
Fit a simple linear regression model with exam as the response and cw as the predictor
```{r}
m1 <- lm(exam ~ cw)
coefficients(m1)
residuals(m1)
```
A student got 4.4 exam marks for each cw mark

calculate the fitted values of the model
```{r}
fitted(m1)
```
plot the line
```{r}
plot(cw, exam, col = "purple", main = "cw vs exam marks", xlab = " coursework mark (out of 20)", ylab = "exam mark (out of 100)")
#plot the fitted line
abline(m1, col = 'black')
```
cwp stores cw mark as if it were out of 10
```{r}
cwp <- cw*5
```
 Fit a simple linear regression model again, using cwp as predictor
```{r}
m2 <- lm(exam ~ cwp)
m2
fitted(m2)
```

```{r}
plot(cwp, exam, col = "purple", main = "cw vs exam marks", xlab = " coursework mark (out of 100)", ylab = "exam mark (out of 100)")
abline(m1, col = 'black')
abline(m2, col = 'purple')
```
There is no difference btw the result we got from rescaling the parameters. The fitted values are identical, and so is the line. Now 1 extra point in te cw (so in reality 1/5 of a point) predicts 0.88 of an extra point in the exam

calculate a prediction for the exam mark achieved by another student who scored 10 marks in the coursework. 
```{r}
predict(m1, newdata = list(cw = 10))
```
My model predicts this student would score 63% in the exam
```{r}
predict(m1, newdata = list(cw = 20))
```
The model predicts the student would score 107% on the exam, which is unreasonable as such a score is impossible.
*why has this happened?*
The model extrapolated, based on the data it had. The model becomes less accurate once we stray away from the range of data we input


Our model suggests that 40 = 19.377 + 4.383 × cw. Rearranging, this gives cw = 4.74, so a
student would need to score at least 5 marks in the coursework to have a predicted exam mark
over 40%.

**2: Haemoglobin data**
```{r}
haem.dat <- read.csv("Haemoglobin.csv")
```

```{r}
haem.dat
```

```{r}
male <- 1 * (haem.dat$gender == 'Male')
male
```
α + weightiβ + e_i        ,i is female
  
α + weightiβ + γ + e_i    ,i is male

the design matrix is:
```{r}
cbind(rep(1,10), "weight" = haem.dat$weight, male)
```
//Xn
i=1
(yi − α − βxi)
2

\\Xn
i=1
(yi − α − βxi)
2

