---
title: "STATISTICAL LABORATORY"
subtitle: "Practical 5"
author: "Dodzia"
output: html_notebook
---

**Poisson Arrival Process**

a)
```{r}
v <- rexp(1000,1)
```
This generates 100 pseudorandom numbers with the exponential distribtion with rate parameter 1

```{r}
vcum <- cumsum(v)
```
this calculates the cumulative sum of the generated numbers stored in v

```{r}
count <- length(which(vcum <= 10))
head(vcum, 25)
which(vcum <= 10)
length(which(vcum <= 10))
```
`which(vcum <= 5)` tells us the indexes of vcum that are less than 10
`length(which(vcum <= 5))` tells us the length of the list of indexed (ie how many indexes)
`count <-` ... assignes this value to an interval. the value of `count` can be understood as the number of events that occur before time t = 10

b)

(i) Adjust the code so that count returns the number of events that occur
before time t = 0.0001. What result do you expect to get for count in this
case? Briefly justify your answer. Run the code a few times to verify your
answer.
```{r}
count <- length(which(vcum <= 0.0001))
```
I expect to get `0` as our rate parameter is 1 (ie we expect the first event to happen around time 1).

(ii) Assume that we want to count the number of events before time t = 1000
and we use the following code:
```{r}
v <- rexp(100,1)
vcum <- cumsum(v)
max(vcum)
count <- length(which(vcum <= 1000))
count
```
It selects all events from v as all happened before time 1000 (we expect the last one to happen around t=100). Instead lets generate `1500` of the iterations as shown below
```{r}
v <- rexp(1500,1)
vcum <- cumsum(v)
max(vcum)
count <- length(which(vcum <= 1000))
count
```
We can see that from 1500 iterations, 1047 happened before time `t=1000`

(c) 
Write a function countevents() that, given a vector v of inter-event times
and a specific time t, returns the number of events that occur before time t.
Note that you need to pay attention to the case where the total of the interevent times may be shorter than the interval. Make sure that your code returns
a warning message when such case occurs. Also, test your function on three
numerical examples; (v1, t1) = ((1, 4, 2, 5, 3), 10), (v2, t2) = ((1, 2, 1), 10) and
(v3, t3) = ((15, 6, 4, 1), 10), and include the output for these examples in your
report.
```{r}
countevents <- function(vector,t) {
  a <- vector[1:length(vector)-1]
  acum <- cumsum(a)
  count <- length(which(acum <= t))
  
  if (max(acum) < t) {
    stop("Caution: the total of the interevent times is shorter than the interval")
  }
  
  return(count)
}
temp1 <- c(1, 4, 2, 5, 3)
temp2 <- c(1, 2, 1)
temp3 <- c(15, 6, 4, 1)
print("temp1")
countevents(temp1,10)
print("temp3")
countevents(temp3,10)
print("temp2")
countevents(temp2,10)
```

(d) For this part, you are asked to simulate a Poisson arrival process by using the
waiting time description given above. This involves simulating inter-event times
using rexp() and using countevents() to determine the number of counts.
To do this, write a function poissonsimulation() which calculates the number
of arrivals until time t in n such trials. The input parameters should be the
exponential parameter λ, the time t, and the number of trials n. The function
should return a vector of length n, where the i
th element of this vector should
correspond to the number of events before time t for the i
th trial. 
For this section, you may assume 0.1 ≤ λ ≤ 1, and 0 < t ≤ 1, otherwise the number
of inter-event times you need to calculate will get too high. Even with this
limitation though you will have to think carefully about how many inter-event
times you need to calculate. 

```{r}
poissonsimulation <- function(n, lambda, t) {
  m <- 1000
  # create empty vector
  result <- c()
  
  # creates m variables with param λ n times
  for (i in 1:n) {
  v <- rexp(m,lambda)
  vcum <- cumsum(v)
  count <- length(which(vcum <= t))
  
  # displays error message if we are selecting all generated values
  if (max(vcum) <= count) {
    print("Caution: the total of the interevent times is shorter than 
          the  interval")
  }
  
  result <- c(result, count)
  }
  # returns result
  return(result)
}

#code from answersheet
poissonsimulation1 <-function(lambda,t,n){
m <- ceiling(100*t)
v <- matrix(0, nrow = m,ncol = n)
counts <- 0
for (i in 1:n){
v[,i] <- rexp(m,lambda)
counts[i] <- countevents(v[,i], t)
}
return(counts)
}

poissonsimulation(10, 0.39, 0.83)
poissonsimulation1(0.39, 0.83, 10)

```

Once complete, poissonsimulation() should return a vector with n elements,
each of which is a count. Run your function for λ = 0.5, t = 1 and n = 50000
to get a sample of n = 50000 counts and find the proportion of each count value
in your sample. How do these proportions compare with the probability mass
function of a Poisson(0.5) distribution?
```{r}
po_sim <- poissonsimulation(n=50000, lambda=0.5, t=1)
```

(f) Explore the distribution of the process you have generated with your simulation
function. Since all the values are discrete, bar plots are better than histograms.
Note you may not only invoke barplot() but also apply table() to the result

```{r}
po_sim_table <- table(po_sim)/50000
po_sim_table
dpois(0:5,lambda = 0.5)
barplot(po_sim_table)
```
The results are aprox equal to the true values of possson distribution


```{r}
po_sim1 <- poissonsimulation(500, 0.1, 0.1)
po_sim_table1 <- table(po_sim1)/500
barplot(po_sim_table, col = 'purple')
```
```{r}
po_sim1 <- poissonsimulation(50000, 0.1, 0.1)
po_sim_table1 <- table(po_sim1)/50000
barplot(po_sim_table, col = 'purple')
```

```{r}
po_sim1 <- poissonsimulation(50000, 0.1, 0.5)
po_sim_table1 <- table(po_sim1)/50000
barplot(po_sim_table, col = 'purple')
```

```{r}
po_sim1 <- poissonsimulation(50000, 1, 0.5)
po_sim_table1 <- table(po_sim1)
barplot(po_sim_table, col = 'purple')
```
The graphs look near identical, but the plotting times differ (the one with n=500 is significantly faster)

2. **Poisson Scatter**
In this exercise we study randomly scattered points on a unit square. This spatial
example is inspired by real world phenomena such as positions of stars on a photographic plate, defects on a sheet of metal, positions of cells on a microscope slide or
raindrops hitting your head.
In many situation it is acceptable to make a few simplifying assumptions: there are
no points occurring more than once, the intensity of points is homogeneous across the
entire square, and that whatever happens in one area does not affect what happens
elsewhere. This leads to a model called Poisson scatter with intensity λ (for some
positive constant λ):

(i) The number N(B) of points in a subset B of the unit square is a Poisson random
variable with mean λ × area(B).

(ii) For disjoint subsets B1, . . . , Bj
, the numbers of points N(B1), . . . , N(Bj ), are
mutually independent.

We now turn our attention to the computational implementation and exploration of
the random scatters described by this model.

(a) Generate 100 independent uniformly distributed points on a unit square and
visualise this in a plot using plot(). You may use the build-in R function
runif() that generates uniformly distributed points on an interval. Repeat this
a few times to get an idea of what such patterns look like.
```{r}
x <- runif(100)
y <- runif(100)
plot(x,y, col='purple', main ="100 (random) points on a unit square", pch=21, cex=0.5)
```

(b) Write a function runifrect() that generates n independent uniformly distributed points on a rectangle [x1, x2] × [y1, y2] within the unity square, that
is, with 0 ≤ x1 < x2 ≤ 1 and 0 ≤ y1 < y2 ≤ 1. The input arguments of this
function should be n and the coordinates of the rectangle. The function should
return a matrix of dimensions n × 2. You may build in checks that activate
warnings to users who plug in coordinates that create non-admissible or degenerated rectangles; ideally by writing a separate function so that this can be used
routinely in the functions you will use in the other parts of this exercise.
```{r}
valid <- function(n, x1, x2, y1, y2) {
  
  if (x2<=x1) {
    print("x2 must be larger than x1")
    return(FALSE)
  }
  if (y2 <= y1) {
    print("y2 must be larger than y1")
    return(FALSE)
  }
  
  if (x2>1 | x1<0) {
    print("x-coordinate out of range")
    return(FALSE)
  }
  
  if (y2>1 | y1<0) {
    print("y-coordinate out of range")
    return(FALSE)
  }
  
  return(TRUE)
}
runifrect <- function(n, x1, x2, y1, y2) {
  x_co <- character(0)
  y_co <- character(0)
  if (valid(n, x1, x2, y1, y2)) {
  x_co <- runif(n, x1, x2)
  y_co <- runif(n, y1, y2)
  
  }
  matrix <- rbind(x_co,y_co)
  print(matrix)
  return(matrix) 
  
}
a <- runifrect(25,0.18, 0.68,0.3,0.58)

```

Write a function plotrectpoints() that visualises your result within the unit
square by tracing the rectangle in red dotted lines and plotting the simulated
points in black. The input arguments of this function should be the coordinates
of the rectangle and a matrix containing the points as its rows. You may find
the function rect() useful; note the arguments border and lty.
```{r}
plotrectpoints <- function(matrix, x1, x2, y1, y2){
  plot(c(),c(), xlim=c(0,1), ylim=c(0,1), xlab = "X", ylab = "Y")
  if (valid(1, x1, x2, y1, y2)) {
    
    rect(x1, y1, x2, y2, border = 2, lty = "dotted", bg="red")
    print(matrix)
    x_coor <- matrix['x_co',]
    y_coor <- matrix['y_co',]
    points(x_coor, y_coor, cex=0.5, pch = 21, bg = "black")
    points(matrix, cex=1.5, pch = 16)
  }
}

plotrectpoints(a, 0,1,0,1)

```
```{r}
plotrectpoints <- function(M, x1, x2, y1, y2){
plot(c(),c(), xlim=c(0,1), ylim=c(0,1), xlab="X", ylab="Y") #empty plot
rect(x1,y1,x2,y2,border=2,lty=3)
points(M, pch=16)
}
plotrectpoints(a, 0,1,0,1)
```

(d) Write a function poissonrect() that generates the realisations of a Poisson
random scatter with positive intensity parameter λ for rectangles. 

The input arguments of this function should be the coordinates of the rectangle and λ. The output value of this function should be a matrix that you can plug into
plotrectpoints(). 

Enjoy your hard work by exploring what such random
pattern looks like. Vary the rectangle sizes and locations and run through a range
of λ. For example, create the corresponding plots for the following combinations:

λ x1 x2 y1 y2
100 0.1 0.5 0.3 0.8
100 0 0.5 0.3 0.4
1000 0 0.5 0.3 0.4
1000 0 0.5 0.3 1

Note that λ corresponds to the expected number of points for the whole unit
square, so you need to expect smaller numbers of points for smaller rectangles,
and for small λ you can easily end up with no points at all.

```{r}
#come back to this me ne comprande pa
poissonrect <- function(lambda, x1, x2, y1, y2) {
    if (valid(lambda, x1, x2, y1, y2)==FALSE) {
      return(FALSE)
    }
   
    x <- rexp(lambda, 0.1)
    xcum <- cumsum(x)
    y <- rexp(lambda, 0.1)
    ycum <- cumsum(y)
    
    matrix <- rbind(xcum,ycum)
    print(matrix)
    return(matrix)
  
}
poissonrect(10, 0.18, 0.68,0.3,0.58)
```

```{r}
#MAKE SURE YOU GET IT
poissonrect <- function(lambda, x1, x2, y1, y2){
if (lambda<=0) stop("Poisson intensity parameter needs to be positive")
if(valid(1,x1,x2,y1,y2)){
  
area <- (x2-x1)*(y2-y1) #can’t be 0 or neg
N <- rpois(1,lambda*area)
M <- matrix(nrow=N, ncol=2)
  #note: number of rows is the random number defined above
M[,1] <- runif(N, min=x1, max=x2)
M[,2] <- runif(N, min=y1, max=y2)
return(M)
}
}
```









RANDOM LECTURE STUFF
```{r}
bs <- rbinom(1000,1, 0.1)
hist(bs)
length(which(bs==1))
length(which(bs==0))
```
```{r}
us <- runif(1000,0,1)
bs2 <- rep(0,1000)
bs2[which(us<0.1)] <- 1
hist(bs2)
```

THE GAMMA DIST
```{r}
invcdf <- function(x, lambda) {
  a <- -(1/lambda)*log(1-x)
  return(a)
}
```




```{r}
Sys.setenv(LANG = "en")
```

