---
title: "Rejection Sampling"
output: html_notebook
---

The function betapdf below is the pdf for the beta distribution with parameters
α, β ∈ Z+.

```{r}
betapdf <- function (x,alpha,beta){
if (x<0){return("Invalid input")}
if (x>1){return("Invalid input")}
if (alpha*beta%%1!=0){return("Invalid input")}
a <-(((x)^(alpha-1))*((1-x)^(beta-1)))
b <-((factorial(alpha-1)*factorial(beta-1))/(factorial(alpha+beta-1)))
d <- a/b
return (d)}
```

plot it for 
```{r}
x <- seq(from=0, to=1, by=0.01 )
```

(a) α = β = 1
```{r}
y <- sapply(x, betapdf, alpha=1, beta=1)
plot(x,y)
```

(b) α = β = 2
```{r}
y <- sapply(x, betapdf, alpha=2, beta=2)
plot(x,y)
```

(c) α = 2, β = 5
```{r}
y <- sapply(x, betapdf, alpha=2, beta=5)
plot(x,y)
```
The function uniform2d below samples 2n values from the U[0, 1] distribution and
arranges them in n pairs

```{r}
uniform2d <- function(n){
a <- runif(n,0,1)
b <- runif(n,0,1)
d <- rbind(a,b)
return (d)}
```

(a) Plot the locations of 10,000 two-dimensional co-ordinates where the x-values and
y-values are both samples from the U[0, 1] distribution.
```{r}
# bad method
plot(c(uniform2d(100)))
```

(b) Modify uniform2d so it generates two-dimensional co-ordinates where the xvalues are samples from 
```{r}
# better method
m <- uniform2d(10000)
plot(m[1,],m[2,])
```
b) U[0, 1] and the y-values are samples from U[0, M] for
some positive constant M.
```{r}
uniform2d <- function(n, M){
  if (M <= 0){return("Invalid input: M should be positive")}
  a <- runif(n,0,1)
  b <- runif(n,0,M)
  d <- rbind(a,b)
  return (d)}
```

```{r}
m <- uniform2d(10000, 5)
plot(m[1,],m[2,])
```

The beta distribution has a finite support, that is, it is only non-zero over a finite interval.
The central idea of the rejection sampling algorithm when sampling from a distribution
with a finite interval is to draw a rectangle around the pdf of that distribution. We then
sample uniformly from the rectangle and reject any co-ordinates that lie inside the rectangle
but do not lie under the curve.

We can use the uniform2d function above to uniformly sample points from the rectangle.

First, however, we need to determine the height of the rectangle in question.
the Beta(2,5) distribution's pdf is
f(x; 2, 5) = = 30x(1 − x)^4

By using optimize find the maximum value of f(x; 2, 5) for x ∈ (0, 1) and thereby show it is in the interval [2, 3].

Let's plot the graph first
```{r}
y <- 30*x*((1-x)**4)
plot(x, y)
```

```{r}
f <- function(y) 30*y*((1-y)**4)
optimize(f,lower=0,upper=1, maximum=TRUE)
```
or
```{r}
Maximum <- optimize(betapdf, alpha=2, beta=5, c(0,1), maximum = TRUE)
Maximum
```

The maximum is around x=0.2, as expected


We now require a function that rejects unsuitable samples from the rectangle [0, 1]×[0, M] (produced by uniform2d).
This will be any co-ordinate (xi, yi) found from the modified uniform2d function for which
yi > f(xi; 2, 5).
```{r}
rejection <- function(A){
  # Code to assign the number of sample values to object n
  n <- ncol(A)
  reject <- NULL
  for (i in 1:n){
    x<-A[1,i]
    # Code to assign the value sampled from [0,M] to object y
    y <- A[2,i]
    if (y > betapdf(x,2,5)){
    reject <- c(reject,i)
    }
  }
  # Code to produce an output from the function
    # returns the accepted values
    return(A[,-reject])
}
```

(b) Uniformly sample 50,000 sample values from the rectangle [0, 1]×[0, M]. Use the
rejection function to reject samples that do not lie under the curve f(x; 2, 5).
Plot the points that have not been rejected in a scatter diagram. What do you
notice?
```{r}
A <- rejection(uniform2d(50000, 2.46))
plot(A[1,],A[2,],xlab = "Support of U[0,1] and Beta[2,5]", ylab = "")
```
or
```{r}
M <- as.numeric(Maximum[2])
A <- uniform2d(50000,M)
B <- rejection(A)
plot(B[1,],B[2,],xlab = "Support of U[0,1] and B[2,5]", ylab = "" )
```

The points appear to be uniformly distributed under a curve that corresponds to the pdf of the Beta(2,5) distribution

(c) Plot the x-values of the points generated in 4(b) in the form of a histogram.
What do you notice?
```{r}
hist(A[1,], breaks = 100, freq = FALSE, xlab="x", col='purple')
```
The frequencies correspond to the pdf of the Beta(2,5) distribution
